# Data Wrangling with R

## Cheat Sheet {-}

  * pipes "%>%" are interpreted as 'and then' in code
    * can be typed or accessed by Ctrl+Alt+M

```{r echo=FALSE,results='asis', eval=FALSE}
library(knitr)
mod1cheatsheet<-read.csv("tables/mod1cheatsheet.csv",header=TRUE)
mod1cheatsheet
kable(mod1cheatsheet,caption = "Data Wrangling Cheat Sheet")
```




## Data Wrangling Using Base R Functions
```{r}
Data<-read.csv("datasets/ClassDataPrevious.csv",header = TRUE)
dim(Data)
```

```{r}
colnames(Data)
```

```{r}
Data[1,2]
```

```{r}
Data[c(1,3,4),c(1,5,8)]
```
To view a column
```{r, eval=FALSE}
Data$Year
Data[,1]
Data[,-c(2:8)]
```

```{r}
which(Data$Sport=="Soccer")
SoccerPeeps<-Data[which(Data$Sport=="Soccer"),]
dim(SoccerPeeps)
```

```{r}
SoccerPeeps_2nd<-Data[which(Data$Sport=="Soccer" & Data$Year=="Second"),]
dim(SoccerPeeps_2nd)
```

```{r}
Sleepy<-Data[which(Data$Sleep>8),]
```

### Changing Column Names

```{r}
names(Data)[c(1,7)]<-c("Yr","Computer")
```

Find and remove missing data

```{r, eval=FALSE}
is.na(Data)
```

```{r}
Data[!complete.cases(Data),]
```
### **Summarizing Variables**

```{r}
apply(Data[,c(2,4,6,8)],2,mean)
```
To not include missing values, use arg na.rm

```{r}
apply(Data[,c(2,4,6,8)],2,mean,na.rm=T)
```
In apply(), the second argument specifies whether to summarize row (put 1) or column (put 2) values

Since some of the means are very high, we can use the median instead to be a little more informative.

```{r}
apply(Data[,c(2,4,6,8)],2,median,na.rm=T)
```
### **Summarizing variable by groups**
use tapply()

```{r}
tapply(Data$Sleep,Data$Yr,median,na.rm=T)
```

```{r}
Data$Yr<-factor(Data$Yr,levels=c("First","Second","Third","Fourth"))
levels(Data$Yr)
```

```{r}
tapply(Data$Sleep,Data$Yr,median,na.rm=T)
```

```{r}
tapply(Data$Sleep,list(Data$Yr,Data$Computer),median,na.rm=T)
```

### **Create a new variable based on existing variable**
```{r}
sleep_mins<-Data$Sleep*60
deprived<-ifelse(Data$Sleep<7,"yes","no")
```

Create catagorical variable based on numerical value

```{r}
CourseLoad<-cut(Data$Courses,breaks=c(-Inf,3,5,Inf),labels=c("light","regular","heavy"))
```

Collapse levels
into upperclassmen and lowerclassmen

```{r}
levels(Data$Yr)
new.levels<-c("und","und","up","up")
Year2<-factor(new.levels[Data$Yr])
levels(Year2)
```
### **Combine data frames**

```{r}
Data<-data.frame(Data,sleep_mins,deprived,CourseLoad,Year2)
head(Data)
```
Can use cbind() alternatively for same result

```{r}
Data2<-cbind(Data,sleep_mins,deprived,CourseLoad,Year2)
```

When combining data frames which have different observations but the same columns, we can merge them using rbind()

```{r}
dat1<-Data[1:3,1:3]
dat3<-Data[6:8,1:3]
res.dat2<-rbind(dat1,dat3)
head(res.dat2)
```
**Export data frame to csv
```{r}
write.csv(Data,file="exports/newdata.csv",row.names=FALSE)
```

### **Sort data frame by column values**
to sort in ascending order by age, then descending
```{r}
Data_by_age<-Data[order(Data$Age),]
Data_by_age_des<-Data[order(-Data$Age),]

```

To sort ascending by age then by sleep:
```{r}
Data_by_age_sleep<-Data[order(Data$Age, Data$Sleep),]
```

***

## Data Wrangling Using dplyr Functions

First we'll clear our environment using rm(list=ls()), then load tidyverse, which contains the dplyr functions:
```{r}
rm(list=ls())
library(tidyverse)
```

```{r}
Data<-read.csv("datasets/ClassDataPrevious.csv",header=TRUE)
```

### Select Specific Columns of a Data Fram

use select() function, two ways to do it

```{r include=FALSE}
select(Data,Year)
```

Or use **Pipes**
```{r include=FALSE}
Data %>% 
  select(Year)
```

Pipes can be typed using either %>% or Ctrl+Shift+M on keyboard. To thing of the operations above, we can read the code as:

1. Take the data frame called Data
2. and then select the column named Year

In this way, we can interpret a pipe as "and then". Commands after pipe should be placed on a new line. Pipes are especially useful for lots of sequential commands.

### Select observations by conditions

The **filter()** function allows us to subset our data based on some conditions, for example, to select students whose favorite sport is soccer:

```{r include=FALSE}
filter(Data,Sport=="Soccer")
```

or use a pipe to store results in a new variable "SoccerPeeps"

```{r}
SoccerPeeps<-Data %>% 
  filter(Sport=="Soccer")

SoccerPeeps_2nd<-Data %>% 
  filter(Sport=="Soccer" & Year=="Second")

Sleepy<-Data %>% 
  filter(Sleep>8)

Sleepy_or_Soccer<-Data %>% 
  filter(Sport=="Soccer"|Sleep>8)
```

### Change Column Name

Changing the names of columns is easy with dplyr, use rename() function

```{r}
Data<-Data %>% 
  rename(Yr=Year,Comp=Computer)
```

### Summarizing Variables

The summarize() function allows us to summarize a column. Suppose we want to find the mean value of the numeric columns: Sleep, Courses, Age, Lunch:

```{r}
Data %>% 
  summarize(mean(Sleep,na.rm=T),mean(Courses),mean(Age),mean(Lunch,na.rm=T))
```

This output is cumbersome, but we can give names to each summary:

```{r}
Data %>% 
  summarize(avgSleep=mean(Sleep,na.rm = T),avgCourse=mean(Courses,na.rm = T),avgAge=mean(Age,na.rm=T),avgLun=mean(Lunch,na.rm = T))
```

As previously seen, some of these variables are suspiciously high, we can use the median instead of mean to get more informative results:

```{r}
Data %>% 
  summarize(avgSleep=median(Sleep,na.rm = T),avgCourse=median(Courses,na.rm = T),avgAge=median(Age,na.rm=T),avgLun=median(Lunch,na.rm = T))
```

### Summarizing Variable by Groups

If we want to find the median amount of sleep for 1st, 2nd, 3rd, and 4th years, we can use the 'group_by()' function.

```{r}
Data %>% 
  group_by(Yr) %>% 
  summarize(medSleep=median(Sleep,na.rm=T))
```

The way we can read the code is:
  1. Get the data frame called Data,
  2. and then group the observations by Yr,
  3. and the find the median amount of sleep by each Yr and store the median in a vector called medSleep
  
The order of the factor levels is in alphabetical, which isn't very useful. We can use the 'mutate()' function whenever we want to transform or create a new variable. In this case, we are transforming the variable 'Yr' by reordering the factor levels with the 'fct_relevel()' function:

```{r}
Data<-Data %>% 
  mutate(Yr<-Yr %>% 
           fct_relevel(c("First","Second","Third","Fourth")))
```

which reads:
  1. Get data frame called 'Data',
  2. and then transform the variable called 'Yr',
  3. and then reorder the factor levels

then we use pipes, the 'group_by()', and 'summarize()' functions like before:

```{r}
Data %>% 
  group_by(Yr) %>% 
  summarize(medSleep=median(Sleep,na.rm=T))
```

This output makes a lot more sense for this context.

To summarize a variable on groups formed by more than one variable, we just add the other variables in the 'group_by()' function:

```{r}
Data %>% 
  group_by(Yr,Comp) %>% 
  summarize(medSleep=median(Sleep,na.rm=T))
```

### Create a New Variable Based on Existing Variable(s)

As mentioned previously, the 'mutate()' function is used to transform a variable or to create a new variable. There are a few variations on this task, based on the type of variable you want to create, and the type of variable it is based on.

**Create a numeric variable based on another numerica variable**

The variable 'Sleep' is in number of hours. Suppose we need to convert the values of 'Sleep' to number of minutes, we can simply perform the following mathematical operation:

```{r}
Data<-Data %>% 
  mutate(Sleep_mins=Sleep*60)
```

and store the transformed variable called 'Sleep_mins' and add 'Sleep_mins' to the data frame called 'Data'.

**Create a binary variable based on a numeric variable**

Create binary variable called 'deprived'. An observation will obtain a value of 'yes' if they sleep less than 7 hours a night, and 'no' otherwise. We will then add this variable to the data frame called 'Data':

```{r}
Data<-Data %>% 
  mutate(deprived=ifelse(Sleep<7,"yes","no"))
```


**Create a categorical variable based on a numeric variable**

Suppose we want to create a categorical variable based on the number of courses a student takes. We will call this new variable 'CourseLoad', which takes on the following variables
  
  * 'light' if 3 courses or less
  * 'regular' if 4 or 5 courses
  * 'heavy' if more than 5 courses
  
and then add 'CourseLoad to the data frame 'Data'. We can use the 'case_when()' function from the 'dplyr' package, instead of the 'cut()' function:

```{r}
Data<-Data %>% 
  mutate(CourseLoad=case_when(Courses<=3 ~ "light",
                              Courses>3 & Courses <=5 ~"regular",
                              Courses>5 ~ "heavy"))
```

Note how the category names are suppied after a specific condition is specified.

**Collapsing Levels**

Sometimes a categorical variable has more levels than needed, so we may want to collapse the levels, as such in this case: collapsing 'Year' into upperclassmen and lowerclassmen.

```{r}
Data<-Data %>% 
  mutate(UpUnder=fct_collapse(Yr,under=c("First","Second"),up=c("Third","Fourth")))
```

here we've created a new variable called 'UpUnder', which is done by collapsing 'First' and 'Second' into a new factor called 'under', and collapsing 'Third' and 'Fourth' into a new factor called 'up'. 'UpUnder' is also added to the dataframe 'Data'.

### Combine Data Frames

To combine data frames which have different observations but the same columns, we can combine them using 'bind_rows()':

```{r}
dat1<-Data[1:3,1:3]
dat3<-Data[6:8,1:3]
res.dat2<-bind_rows(dat1,dat3)
head(res.dat2)
```

'bind_rows()' works the same way as 'rbind()'. Likewise, we can use 'bind_cols()' instead of 'cbind()'.

### Sort data frame by column values

To sort your data frame in ascending order by 'Age':

```{r}
Data_by_age<-Data %>% 
  arrange(Age)
```

To sort in descending order by 'Age':

```{r}
Data_by_age_des<-Data %>% 
  arrange(desc(Age))
```

To sort in ascending order by 'Age' first, then 'Sleep':

```{r}
Data_by_age_sleep<-Data %>% 
  arrange(Age,Sleep)
```

### More About Combining Datasets

```{r}
rm(list = ls())
#################
##load packages##
#################

library(nycflights13)
library(tidyverse)

##see dataframes from packages
##View(flights)

##check documentation
##?flights

##View(airlines)
##?airlines

##merge data frames that share one column with the same name
flight_airlines<-flights%>%
  inner_join(airlines,by="carrier")

##merge data frames with multiple shared common columns
##View(weather)
flights_weather<-flights%>%
  inner_join(weather, by=c("year","month","day","hour","origin"))

##merge data frames when columns have different names but same content
##View(airports)
flights_airports<-flights%>%
  inner_join(airports,by=c("dest"="faa"))

##similar function in base R, merge()

flight_airlines2<-merge(flights,airlines, by="carrier")
##View(flight_airlines2)

flights_weather2<-merge(flights,weather, by=c("year","month","day","hour","origin"))
##View(flights_weather2)

##not sure if you notice, the merge() function takes longer to run. 
##use proc.time() to time how long your code takes to run

begin<-proc.time()
flight_airlines2<-merge(flights,airlines, by="carrier")
proc.time()-begin

begin<-proc.time()
flight_airlines<-flights%>%
  inner_join(airlines,by="carrier")
proc.time()-begin

```

Basically, use innerjoin rather than merge because it's an order of magnitude faster

### A Note About Missing Values

Let's go over some standard missing values that R recognizes and how to handle nonstandard ones that R will not recognize

```{r}
rm(list = ls())
library(tidyverse)
Data<-read.csv("datasets/missing.csv",header=TRUE)
Data
```

Some of these observations are missing values

We can apply is.na() to dataframe to see which entries are viewed as missing:

```{r}
is.na(Data)
```

As we can see, R only recognized entries with NA and NaN as missing. These are the standard values for missing entries. Any other way is not recognized.

**Note**:

  * NaN represents undefined number
  * NA represents missing value

We can convert the non standard missing values to \'standard missing values\' using the 'replace()' function within the 'mutate()' function:

```{r}
Data<-Data %>% 
  mutate(Weight = replace(Weight, Weight == "na", NA))%>%
  mutate(Weight = replace(Weight, Weight == "N/A", NA))%>%
  mutate(Weight = replace(Weight, Weight == "", NA))

is.na(Data)
```

And just like that, the rest of the missing values are recognized as missing as they should be.



