# Multiple Linear Regression

## A Note on Notes

I think the note taking for these courses has become inefficient; so from now on I'll just be reviewing prof's notes and then doing the R Tutorial.

So, without further adieu...

## R Tutorial

We'll be using data on cherry trees from `openintro`.

```{r}
rm(list=ls())
library(openintro)
Data<-openintro::cherry
```

## Scatterplot matrix

We'll use the `ggpairs()` function from `GGally` package:

```{r}
library(GGally)
GGally::ggpairs(Data)
```

* the off diagonals give us the scatterplot and correlation between corresponding pairs of variables.

## Fit MLR Using `lm()`

To fit multiple linear regression (MLR)

```{r}
result<-lm(volume~diam+height,data=Data)
```

Another way to do this is:

```{r}
result<-lm(volume~.,data=Data)
```

The `.` informs `lm()` to use every column other than `volume` in the data frame as predictors.

```{r}
summary(result)
```

so the estimated regression equation is $\hat{y}=-57.9877+4.7802\text{(diam)}+0.3393\text{(height)}$

## Inference with MLR

Since the anova F statistic is 255 with small p-value, data supports that our model is useful.

The confidence intervals for the coefficients can be found using `confint()`:

```{r}
confint(result,level = 0.95)
```

Confidence interval for the mean response and prediction interval for a new observation given a specific value of the predictors can also be found using `predict()` e.g.

```{r}
newdata<-data.frame(diam=10,height=80)

predict(result,newdata,level=0.95,interval = "confidence")
```

```{r}
predict(result,newdata,level=0.95,interval = "prediction")
```

These are the same functions as we did in SLR.

Assessing regression assumptions:

```{r}
ggplot(Data,aes(x=diam,y=volume))+
  geom_point()+
  geom_smooth(method = lm)
```

```{r}
ggplot(Data,aes(x=height,y=volume))+
  geom_point()+
  geom_smooth(method = lm)
```

```{r}
library(MASS)

par(mfrow=c(2,2))
plot(result)
```

























