# Basics of R

## Getting Started with R

### Question 1

**(a)**
```{r}
cars.df<-mtcars
```

**(b)** 
According to the environment window, there are 32 observations of 11 variables in the dataset *mtcars*.

### Question 2

**(a)**
```{r}
students.df<-read.table("datasets/students.txt",header=TRUE)
```

**(b)**
According to the environment window, there are 249 observations of 9 variables in the dataset *students.txt*.

### Question 3
**(a) - (h)**

The packages *tidyverse, faraway, MASS, leaps, ROCR, nycflights13, gapminder, palmerpenguins* were installed.

### Question 4

```{r}
library(faraway)
corn.df<-cornnit
```

## Topic B.3: Data Types & Structures in R

### Question 5

**(a)** 2020_Major    Valid

**(b)** .2020.Age   Invalid, number follows .

**(c)** #Courses.2020   Invalid, # not allowed

**(d)** _courses_2020   Invalid, cannot start with underscore

**(e)** Fav_Sport20   Valid

**(f)** major 2020    Invalid, space not allowed

**(g)** age(2020)   Invalid, parentheses not allowed

**(h)** FavSport_2020   Valid

### Question 6

```{r}
practice<-c(13,91,36,95,9,3,61,20,22,97)
class(practice)
```
### Question 7

**(a)** practice[5]==5    False
```{r}
practice[5]==5
```

**(b)** practice[10]!=97    False
```{r}
practice[10]!=97 
```

**(c)** (practice[1]+practice[2])<104    False
```{r}
(practice[1]+practice[2])<104
```

**(d)** (practice[1]+practice[2])<=104   True
```{r}
(practice[1]+practice[2])<=104
```

**(e)** (practice[2]==91) & (practice[9]==22)   True * True=True
```{r}
(practice[2]==91) & (practice[9]==22)
```

**(f)** (practice[5]<9) | (practice[6]>=4)    False + False = False
```{r}
(practice[5]<9) | (practice[6]>=4)
```

### Question 8

```{r}
Mat.A<-matrix(c(4,6,1,2,3,1),nrow = 2,ncol = 3)

Mat.A
```

**(a)**
```{r}
colnames(Mat.A)<-c("Huey","Dewey","Louie")

Mat.A
```
**(b)** Output of Mat.A[2,1] would be 6.
```{r}
Mat.A[2,1]
```

**(c)** Output of dim(Mat.A) would be (2,3).
```{r}
dim(Mat.A)
```

### Question 9
```{r}
factor(practice)
```

The order of the levels in the factor *practice* are:
3 9 13 20 22 36 61 91 95 97

## R Markdown

### Question 10
As evidenced by the above, my answers were typed up using R Markdown, and an HTML file was created.